*Prochlorococcus marinus* responses to light and oxygen

### 2024-02-26 Code/Manuscript_O2.Rmd
text edits

Low-Light clades thrive in deeper ocean waters, extending beyond 200 m in depth [@partenskyProchlorococcusMarinePhotosynthetic1999], where only ~1% of the surface irradiance penetrates, primarily in the blue (450 nm) to green (520 nm) spectral range [@holtropVibrationalModesWater2021]. Clade LLIV, including cultured strain MIT9313, falls near the base of the *Prochlorococcus* radiation, and is characterized by preference for low light, typical of depths from 120 m to 200 m [@partenskyProchlorococcusMarinePhotosynthetic1999]. LLIV members are, as yet, the only cultured strains to have been found in Oxygen Minimum Zones (OMZ).  Some as yet uncultured *P. marinus* strains in clades LLV and LLVI also thrive in OMZ of the subtropical Atlantic and Pacific Oceans, where dissolved oxygen concentrations [O~2~] can be less than 20 µM [@ulloaCyanobacteriumProchlorococcusHas2021; @lavinNovelLineagesProchlorococcus2010; @goerickeNovelNicheProchlorococcus2000; @garcia-robledoCrypticOxygenCycling2017; @johnsonEnergeticsGrowthKinetics1999]. *Prochlorococcus marinus* LL ecotypes may indeed dominate the phytoplankton within OMZ
[@lavinNovelLineagesProchlorococcus2010;
@ulloaCyanobacteriumProchlorococcusHas2021; @partenskyProchlorococcusAdvantagesLimits2010], and may be net O~2~ consumers [@partenskyComparisonPhotosyntheticPerformances2018]. Clades LLII and LLIII are grouped together as the second oldest phylogenetic lineages diversifying in the *P. marinus* radiation, and share an affinity for low light; LLIII cultured strain SS120, was isolated from 135 m depth in the Sargasso Sea. Clade LLI  includes cultured strain NATL2A, which prefers moderate irradiance typically between 30 and 100 m depth.

HL clades are the most recently evolved lineages, with reduced genome sizes in comparison to LL ecotypes. High-Light clades are typically dominant picophytoplankters in near-surface, oligotrophic waters, characterized by high light levels. Clade HLI, including cultured strain MED4,is adapted to higher iron, and lower temperatures, and originated from the Mediterranean Sea, just 5 m below the surface [@partenskyProchlorococcusMarinePhotosynthetic1999].  Clade HLII adapted to higher iron, and higher temperatures, is the most abundant *P. marinus* clade in the North Atlantic and North Pacific Oceans, often constituting over 90% of the total population [@partenskyProchlorococcusMarinePhotosynthetic1999],  and are most numerous around 50 m depth [@partenskyProchlorococcusMarinePhotosynthetic1999]. Clade HLIII/IV is adapted to lower iron [@kentParallelPhylogeographyProchlorococcus2019; @johnsonNichePartitioningProchlorococcus2006; @zinserInfluenceLightTemperature2007].

@@

Near the equator, photoperiod remains relatively constant at the ocean surface, approximately 12 hours (h) of daylight and 12 h of darkness throughout the year. The effective length of the photoperiod does, however attentuate with depth as dawn and dusk light at depth drops below the level needed for biological processes. As *P. marinus* potentially expands its temperature-permissive niches poleward into temperate regions  [@flombaumPresentFutureGlobal2013; @bartonAnthropogenicClimateChange2016], it will encounter more
pronounced seasonal variations in photoperiod regimes both at surface and at depth, with potentially complex effects upon viable growth niches [@liDiatomGrowthResponses2017; @prezelinDielPeriodicityPhytoplankton1992]. For example, Vaulot *et al*. (1995) @vaulotGrowthProchlorococcusPhotosynthetic1995 showed that *Prochlorococcus* replication of DNA occurs in the afternoon, while cell division occurs at night.  To our knowledge, no study has as yet addressed *P. marinus* growth responses in relation to a  range of photoperiods.

>>>>

Low-Light clades thrive in deeper ocean waters, extending beyond 200 m in depth [@partenskyProchlorococcusMarinePhotosynthetic1999], where only ~1% of the surface irradiance penetrates, primarily in the blue (450 nm) to green (520 nm) spectral range [@holtropVibrationalModesWater2021]. Clade LLIV, including cultured strain MIT9313, falls near the base of the *Prochlorococcus* radiation, and is characterized by preference for low light, typical of depths from 120 m to 200 m [@partenskyProchlorococcusMarinePhotosynthetic1999]. LLIV members are, as yet, the only cultured strains to have been found in Oxygen Minimum Zones (OMZ).  Some as yet uncultured *P. marinus* strains in clades LLV and LLVI also thrive in OMZ of the subtropical Atlantic and Pacific Oceans, where dissolved oxygen concentrations [O~2~] can be less than 20 µM [@ulloaCyanobacteriumProchlorococcusHas2021; @lavinNovelLineagesProchlorococcus2010; @goerickeNovelNicheProchlorococcus2000; @garcia-robledoCrypticOxygenCycling2017; @johnsonEnergeticsGrowthKinetics1999]. *Prochlorococcus marinus* LL ecotypes may indeed dominate the phytoplankton within OMZ
[@lavinNovelLineagesProchlorococcus2010;
@ulloaCyanobacteriumProchlorococcusHas2021; @partenskyProchlorococcusAdvantagesLimits2010], and may be net O~2~ consumers [@partenskyComparisonPhotosyntheticPerformances2018]. Clade LLI includes cultured strain NATL2A, which prefers moderate irradiance typically between 30 and 100 m depth. Clades LLII and LLIII are grouped together as the second oldest phylogenetic lineages diversifying in the *P. marinus* radiation, and share an affinity for low light.

HL clades are the most recently evolved lineages, with reduced genome sizes in comparison to LL ecotypes. High-Light clades are typically dominant picophytoplankters in near-surface, oligotrophic waters, characterized by high light levels. Clade HLI is adapted to higher iron, and lower temperatures, and originated from the Mediterranean Sea, just 5 m below the surface [@partenskyProchlorococcusMarinePhotosynthetic1999].  Clade HLII adapted to higher iron, and higher temperatures, is the most abundant *P. marinus* clade in the North Atlantic and North Pacific Oceans, often constituting over 90% of the total population [@partenskyProchlorococcusMarinePhotosynthetic1999],  and are most numerous around 50 m depth [@partenskyProchlorococcusMarinePhotosynthetic1999]. Clade HLIII/IV is adapted to lower iron [@kentParallelPhylogeographyProchlorococcus2019; @johnsonNichePartitioningProchlorococcus2006; @zinserInfluenceLightTemperature2007].

@@

Near the equator, photoperiod remains relatively constant at the ocean surface, approximately 12 hours (h) of daylight and 12 h of darkness throughout the year. The effective length of the photoperiod does, however attenuate with depth as dawn and dusk light at depth drops below the level needed for biological processes. As *P. marinus* potentially expands its temperature-permissive niches poleward into temperate regions  [@flombaumPresentFutureGlobal2013; @bartonAnthropogenicClimateChange2016], it will encounter more
pronounced seasonal variations in photoperiod regimes both at surface and at depth, with potentially complex effects upon viable growth niches [@liDiatomGrowthResponses2017; @prezelinDielPeriodicityPhytoplankton1992]. For example, Vaulot *et al*. (1995) @vaulotGrowthProchlorococcusPhotosynthetic1995 showed that *Prochlorococcus* replication of DNA occurs in the afternoon, while cell division occurs at night.  To our knowledge, no study has as yet addressed *P. marinus* growth responses in relation to a  range of photoperiods.
